Transport map accelerated-PAIS, and application to inverse problems arising from multiscale stochastic reaction networks

### 2019-11-01 Resubmission/TPAIS_Sept_17.tex
Resubmission folder and response.tex

@@

So far we have seen that the PAIS algorithm is efficient for low dimensional posterior sampling.
PAIS is particularly useful if the posterior is multimodal, or inconveniently shaped such as the
Rosenbrock density discussed in Chapter~\ref{sec:LPAIS}. However, we have had trouble when
attempting to increase the dimension of the parameter space. This was seen when we attempted to
recover the initial condition for the wave equation in Chapter~\ref{sec:PAIS}.

In this Chapter we look at a method which will allow us to improve efficiency in our proposal distributions. The ideal sampling
situation in MCMC is to explore an $d$-dimensional Gaussian distribution (or other standard distribution) using a near proportional
proposal distribution. We therefore aim to transform an arbitrary distribution, $\mu_\theta$, in to a Gaussian distribution, which we can then sample from efficiently, and finally transform this sample back into a sample from $\mu_\theta$. In this way we hope to be able to sample from any distribution $\mu_\theta$ with a similar efficiency as we obtain when sampling from a Gaussian distribution. In particular for the PAIS algorithm, resampling schemes are more efficient in higher
dimensions if the dimensions are uncorrelated. The work of Moselhy and
Marzouk~\cite{el2012bayesian}, Parno and Marzouk~\cite{parno2014transport} and
Parno~\cite{parno2015transport} provides a method of constructing such a map as part of an adaptive
Metropolis-Hastings algorithm.

>>>>

\section{Deterministic Resampling Methods}\label{sec:intro_optim}

Optimal transportation arises in many industries. It deals with the efficient allocation
of a particular resource to someone who wishes to use this resource. For example a fresh food
supplier may have many depots around the country, each with a certain number of vans capable of
carrying a fixed amount of product a particular distance in a day. The supplier may need to supply
many supermarkets up and down the country with their produce. The supplier naturally wishes to
perform this task quickly, and using as little fuel as possible. To achieve this goal the supplier
can choose which depot or depots deliver to which supermarket, and whether a van stops at zero, one
or more supermarkets on its route. The efficient solution of this problem can make a huge difference
to the profits of the supplier.

The supplier knows the initial distribution, $\mu$, where we have all of the fresh produce,
and they also know where we want it to go, the target distribution, $\eta$. We need to find a map which converts
the initial distribution, $\mu$, into the target, $\eta$. These
maps are known as transport maps. We can also contrain the problem to one where we must use as little fuel as possible, the map satisfying this condition is known as the optimal transport map.

\begin{dfn}[Transport Map~\cite{reich2013ensemble}]\label{def:transport_map}
Given a random variable $X \sim \mu$, $X \in \mathcal{X}$, and a second probability measure $\nu$, a
measurable function $T\colon \mathcal{X}\rightarrow\mathcal{X}$ is a transport map if the induced
random variable $Y = T(X)$ satisfies
\[
	\int_\mathcal{X} \! f(y)\nu(\text{d}y) = \int_\mathcal{X} \! f(T(x))\mu(\text{d}x),
\]
for all functions $f\colon \mathcal{X} \rightarrow \mathbb{R}$.
\end{dfn}

Mathematically, the formulation of this assignment problem is quite abstract. The original
formulation by Monge in 1781~\cite{monge1781memoire}, is to find the transport map, $T(\cdot)$,
which minimises the integral
\[
	\int_\mathcal{X} \! c(s, T(s)) \, \text{d}\mu(s),
\]
subject to the constraint that $T(\mu) = \nu$. We are able to choose what we consider to be
expensive by choosing the cost function $c(x,y)$ which is non-negative for all states
$x,y\in\mathcal{X}$.

A more widely used formulation of the Monge problem was presented by Kantorovitch in
1942~\cite{kantorovitch1958translocation}. This formulation considers a space of joint probability
distributions known as the space of transference plans, $\Pi(\mu, \nu)$.

\begin{dfn}[Coupling, Transference Plan~\cite{reich2013ensemble}]\label{def:coupling}
Given two distributions $\mu$ and $\nu$ on a space $\mathcal{X}$, a coupling of these distributions
is a pair $Z = (X,Y)$ such that $X\sim\mu$ and $Y\sim\nu$. The joint measure $\eta$, where
$Z\sim\eta$, on $\mathcal{Z}=\mathcal{X}\times\mathcal{X}$ is known as the transference plan for
this coupling.
\end{dfn}

The optimisation problem then becomes
\[
	\int_{\mathcal{X}\times\mathcal{X}} \! c(x, y) \, \text{d}\gamma(x, y),
\]
subject to $\gamma(x,y) \in \Pi(\mu, \nu)$. Kantorovitch also developed the field of
linear programming to help deal with solutions to this
problem~\cite{kantorovich1960mathematical,kantorovitch1958translocation}.

When the cost function is $c(x,y) = \|x - y\|^2$, if $\mu$ is absolutely continuous with respect to
Lebesgue measure, and both $\mu$ and $\nu$ have finite second order moments, then it is guaranteed
that there is a unique optimal transference plan, between $\mu$ and $\nu$~\cite{knott1984optimal}.

%%%%%%%%%%%%

\subsection{Coupling Probability Distributions}\label{sec:coupling}

In this section we return to the idea of coupling probability distributions, and motivate their use
in resampling algorithms. We defined a coupling in Definition~\ref{def:coupling} as a pair of random
variables $Z = (X,Y)$, the joint probability distribution for this pair being a quantity of
interest. In this section we assume that the marginal distributions of both $X$ and $Y$ are known,
later on we relax this so that one of these distributions is not known but can be approximated by a
sample. All distributions used in the later chapters will be absolutely continuous on $\mathcal{X} =
\mathbb{R}^n$, so we have $\mu(\text{d}x) = \pi_0(x)\text{d}x$ and $\nu(\text{d}x) =
\pi(x)\text{d}x$. Having absolutely continuous marginals does not mean that the coupling itself is
absolutely continuous.

\begin{figure}
\centering
\subfigure[$\rho = 0.1$]{\includegraphics[width=0.45\textwidth]{images/coupling_indep}}
\subfigure[$\rho = 0.95$]{\includegraphics[width=0.45\textwidth]{images/coupling_correlated}}
\caption{Couplings with different correlations.}
\label{fig:cts_couplings}
\end{figure}

We now look at couplings between continuous random variables. As an example we consider coupling two
univariate Gaussian distributions, $X_1 \sim \mathcal{N}(\mu_1, \sigma^2_1)$ and $X_2 \sim
\mathcal{N}(\mu_2, \sigma^2_2)$. We know that the transference plan $\gamma(x_1, x_2)$ will be the
bivariate Gaussian distribution,
\[
	Z = (X_1, X_2), \quad Z \sim \mathcal{N}\left(\begin{pmatrix}\mu_1\\\mu_2\end{pmatrix},
		\begin{pmatrix} \sigma^2_1 & \rho\sigma_1\sigma_2 \\ \rho\sigma_1\sigma_2 & \sigma^2_2
		\end{pmatrix}\right).
\]
The form of this bivariate Gaussian is restricted by the marginal distributions to have this
covariance matrix, where we can choose the correlation $-1 \leq \rho \leq 1$. This unknown
correlation means that there are infinitely many couplings between $X_1$ and $X_2$, some examples
are given in Figure~\ref{fig:cts_couplings}, which leads us back to the idea of optimising the
transport map to maximise some criterion. We might want to maximise the correlation between the two
random variables. If we consider the conditional distribution $X_1|X_2 \sim \mathcal{N}(x_1; \mu,
\sigma^2)$, where
\[
	\mu = \mu_1 - \frac{\rho\sigma_1\sigma_2}{\sigma^2_2}(\mu_2 - x_2) \quad \text{and} \quad \sigma^2 =
		\sigma^2_1 - \frac{(\rho\sigma_1\sigma_2)^2}{\sigma_2^2} = (1-\rho^2)\sigma_1^2,
\]
we notice that as $\rho\rightarrow 1$ the variance $\sigma^2 \rightarrow 0$, and we obtain a
deterministic coupling with transport map $T\colon x_1 \mapsto \mu_1 -
\frac{\sigma_1}{\sigma^2}(\mu_2 - x_2)$. If we set $\mathcal{N}(\mu_2, \sigma^2_2) = \mathcal{N}(0,
1)$ we see that this coupling matches our standard map for transforming standard Gaussian random
variables into general Gaussian random variables.

Here we consider an optimal coupling to be one which maximises the correlation between $X_1$ and
$X_2$. Maximising the correlation is equivalent to minimising the expectation of $(X_2-X_1)^2$,
\[
	\mathbb{E}[(X_2-X_1)^2] = \mathbb{E}[X_1^2]+\mathbb{E}[X_2^2] - 2\mathbb{E}[X_1]\mathbb{E}[X_2]-
		2\text{cov}(X_1, X_2).
\]
The coupling which maximises the correlation is therefore the solution to the Monge-Kantorovich
problem with cost function
\[
	c(X_1,X_2) = \|X_1-X_2\|^2.
\]
This problem always has a solution since for any marginals densities $\pi_X(x)$ and $\pi_Y(y)$, the
trivial coupling $\pi_Z(x,y) = \pi_X(x)\pi_Y(y)$ exists.

We now look at couplings between two discrete random variables. Unlike in the continuous case where
the coupling had support over all of $\mathcal{X}\times\mathcal{X}$, for couplings between discrete
random variables this is rarely the case.

Consider two discrete R.V.s
\[
	Y_1 \sim \mu_{Y_1} \quad \text{and} \quad Y_2 \sim \mu_{Y_2},
\]
defined over a parameter space $\mathcal{Y} = \left\{a_1, \dots, a_M\right\}$, $a_i \in \mathbb{R}$.
Assign the mass functions
\[
	\mathbb{P}(Y_1 = a_i) = \frac{1}{M} \quad \text{and} \quad \mathbb{P}(Y_2 = a_i) = w_i,
\]
where $w_i \geq 0$ and $\sum_i \! w_i = 1$. All possible couplings $Z = (Y_1, Y_2)$ can be described
by the matrix $T \in \mathbb{R}^{M\times M}_+$, such that
\begin{equation}\label{eq:coupling_conditions}
\mathbb{P}(Y_2) = MT\mathbb{P}(Y_1), \quad \text{subject to} \quad \sum\limits_{i=1}^M \! t_{ij} =
\frac{1}{M}, \quad \sum\limits_{j=1}^M \! t_{ij} = w_i,
\end{equation}
where $[T]_{ij} = t_{ij}$.

We present two special cases, first the map with zero correlation, or trivial map,
\[
T^t = \frac{1}{M}\begin{bmatrix} w_1 & \dots & w_1 \\ \vdots & & \vdots \\ w_M & \dots & w_M
\end{bmatrix}
\]
It is clear that, after renormalisation, along any row we obtain a uniform distribution,
$\mathbb{P}(Y_1)$, and along any column we obtain the distribution $\sum_{i=1}^M
\delta_{a_i}(\cdot)w_i$, $\mathbb{P}(Y_2)$. Hence our conditionals are equal to our marginals and
there is no correlation.

The second case is more interesting. The coupling which maximises the correlation between the two
marginals, $T^*$, is the solution to the optimisation problem
\[
	T^*  = \arg\min_T \! \sum\limits_{i,j=1}^M \! t_{ij}(a_i - a_j)^2,
\]
subject to the row and column constraints on $T$ in Equation~\eqref{eq:coupling_conditions}~\cite{reich2015probabilistic}.

We now look at a simple example of a discrete coupling. Given a state space $\mathcal{Y} = \{a_1,
\dots, a_M\}$ where $M=5$, and two probability distributions defined over $\mathcal{Y}$,
\[
\mathbb{P}(Y_1 = a_i) = \frac{1}{M} \quad \text{and} \quad \mathbb{P}(Y_2 = \mathbf{a}) = (0.1,
0.25, 0.05, 0.3, 0.3)^\top.
\]
If we assume that the states $a_i$ are sorted so that $a_i < a_{i+1}$ for all $i=1,\dots,M-1$, then
the optimal coupling is
\[
	T^* = \begin{bmatrix} 0.1 & 0      & 0    & 0    & 0 \\
					       0.1 & 0.15 & 0    & 0    & 0 \\
					       0    & 0.05 & 0    & 0    & 0 \\
					       0    & 0      & 0.2 & 0.1 & 0 \\
					       0    & 0      & 0    & 0.1 & 0.2
		\end{bmatrix}.
\]
For an ordered state space we can construct this coupling following this staircase pattern by
starting in the top left and filling first the row then the column heading towards the bottom right
corner. We notice that this staircase structure corresponds to the shape seen in the highly
correlated coupling in Figure~\ref{fig:cts_couplings} (b). The $2M-1$ constraints mean that we have
at most $2M-1$ non-zero entries, so the support of the joint distribution is much smaller than the
full space.

The product $MT^*$ results in a column stochastic matrix $P$. If we want to draw an evenly weighted
sample, $\{s_i\}_{i=1}^5$, which obeys the probability law $\mathbb{P}(Y_2)$, we can use the columns
of $P$ as probability distributions,
\[
	s_1 \sim \frac{1}{2}\delta_{a_1}(\cdot)+\frac{1}{2}\delta_{a_2}(\cdot), \quad s_2 \sim
		\frac{3}{4}\delta_{a_2}(\cdot)+\frac{1}{4}\delta_{a_3}(\cdot), \quad s_4 \sim
		\frac{1}{2}\delta_{a_4}(\cdot)+\frac{1}{2}\delta_{a_5}(\cdot),
\]
and deterministically $s_3 = a_4$, $s_5 = a_5$. What we have done here is used an optimal coupling
between weighted and unweighted random variables to resample from the weighted distribution.

%%%%%%%%%%%%

\subsection{Resampling with Filters}
\label{sec:OT}

Resampling is a method which is distinct from most of the Statistical literature in that rather than
sampling from a known parametric distribution which is usually the aim, we instead attempt to sample
from empirical distributions. These distributions often arise from a sample, $\{(w_i,
\theta_i)\}_{i=1}^n$, in the form,
\begin{equation}\label{eq:empirical_dist}
\pi(\cdot) = \sum\limits_{i=1}^n \! w_i\delta_{\theta_i}(\cdot), \quad \text{where}
\quad\sum\limits_{i=1}^n \! w_i = 1.
\end{equation}
The weights, $w_i$ may be equal to $1/n$ or they may all be different.
\begin{dfn}[Resampling]\label{def:resampling}
Given a discrete empirical distribution of the type in Equation~\eqref{eq:empirical_dist}. Resampling provides
us a new discrete distribution of the form
\[
	\tilde{\pi}(\cdot) = \frac{1}{n}\sum\limits_{i=1}^n \! n_i\delta_{\theta_i}(\cdot), \quad
		\text{where} \quad n_i \in \{0,\dots,n\}, \quad \sum\limits_{i=1}^n \! n_i = n.
\]
The density $\tilde{\pi}$ is in some way close to the original density $\pi$.
\end{dfn}
In Definition~\ref{def:resampling} the condition on the sum of the $n_i$ restricts $\tilde{\pi}$ to
be composed of the same number of samples as $\pi$. This is not necessary although as mentioned in
Section~\ref{sec:gmh} one can run into trouble when this is not the case. These problems arise
because the empirical density $\pi$ is usually approximating some continuous distribution, $f$, and
over reliance on one approximation will cause the tails of the $f$ to be underestimated. If a sample
is required, one uses $n_i$ lots of $\theta_i$, otherwise statistics can be computed directly from
$\tilde{\pi}$.

%%%%%%

\subsubsection{Bootstrap Resampling}

One of the first resamplers capable of accomplishing the feat in Definition~\ref{def:resampling} was
the bootstrap resampler (BSR)~\cite{efron1992bootstrap}.

\begin{table}[htpb]
\begin{algorithm}[H]
\DontPrintSemicolon
\BlankLine
Obtain sample, $\{x_i\}_{i=1}^n$ from some distribution.\;
Construct
\[
	\hat{F} = \sum\limits_{i=1}^n \! n_i\delta_{x_i}(\cdot), \quad n_i = \sum\limits_{k=1}^n \!
		\text{I}(x_i = x_k).
\]

Sample with replacement $X^* = [x^*_1,\dots,x_n^*]^\top \sim \hat{F}$. i.e. $X^* \sim
\text{Multi}(1, \frac{1}{n}[n_1,\dots,n_n]^\top)$.\;
\caption{The Bootstrap Resampler~\cite{efron1992bootstrap}.\label{alg:bs_resampler}}
\end{algorithm}
\end{table}

Algorithm~\ref{alg:bs_resampler} presents the BSR when we have an evenly weighted sample, this can
just as easily be extended to a weighted sample, $\{(w_i, x_i)\}_{i=1}^n$. In this case the
probability vector in the Multinomial distribution becomes $\tilde{\mathbf{w}} = [\tilde{w}_1,
\dots, \tilde{w}_n]^\top$ where $\tilde{w}_i = \sum_{k=1}^n \! w_i\text{I}(x_i = x_k)$.

%%%%%%

\subsubsection{The Bootstrap Particle Filter}

The data assimilation community is concerned with predicting the state of time dependent systems
given noisy observations. This type of problem is known as the filtering problem. Usually there is a
model for the system although it may be expensive to evaluate. We assume that we have a prediction,
$x(t)=[x_1(t),\dots,x_N(t)]^\top$, from the model calculated using observations, $D_i(t-1)$. We also
observe some data at time $t$, $D_i(t)$. An application of Bayes' formula allows us to combine these
two pieces of information to refine the prediction.
\begin{dfn}[Particle Filter]
A particle filter is an algorithm which solves the filtering problem.
\end{dfn}

\begin{table}[htpb]
\begin{algorithm}[H]
\DontPrintSemicolon
\BlankLine
Define a system model,
\[
	x_i(t) = f_{t-1}(x_i(t-1), \xi_i(t-1)), \quad \text{where} \quad f_t\colon
		\mathbb{R}^n\times\mathbb{R}^m\rightarrow\mathbb{R}^n, \quad \xi_i(t-1) \sim \mathcal{N}(0,
\sigma^2).
\]

\nonl \textbf{Predict}\;
Sample $\xi_i(t-1) \sim \mathcal{N}(0, \sigma^2)$.\;
Calculate $x_i^*(t) = f_{t-1}(x_i(t-1), \xi_i(t-1))$.\;
\nonl\;
\nonl \textbf{Update}\;
For each particle calculate
\[
	w_i = \frac{\ell(D_i(t)| x_i^*(t))}{\sum_{k=1}^N \! \ell(D_k(t)| x_k^*(t))}.
\]

Resample $x(t) \sim \text{Multi}(1, w/\|w\|_1)$, where $w = (w_1,\dots,w_N)^\top$.
\caption{The Bootstrap Particle Filter~\cite{gordon1993novel}.\label{alg:bs_filter}}
\end{algorithm}
\end{table}
The bootstrap filter is an early particle filter, consisting of two steps. The first is to propogate
the prior's predictions through the system model. We then assign a weight to each particle based on
how likely the data we have just received is given the prior prediction. Finally we use the
bootstrap resampler to convert these weighted particles into the estimate of the current state. This
current state can become the prior information for the next iteration.

%%%%%%

\subsubsection{Linear Ensemble Transform Filters}

The bootstrap filter is a random transformation of the empirical density, $\pi$, and the transformed
density, $\tilde{\pi}$, has the same support as $\pi$, with a number of copies of each
particle. It is often undesirable for us to have copies since this means that we do not cover as
much of the sample space as we could. This leads to wasted computation, poorer quality samples and a
worse approximation of the densities of interest. One method of reducing the number of repeats in a
sample is to allow $\tilde{\pi}$ to be built from linear combinations of the original sample. This
means that the support of the density $\tilde{\pi}$ is not restricted to being the same as $\pi$ and
reduces the likelihood of having two particles in the same position. Particle filters which use
linear combinations to produce the transformed density are known as Linear Ensemble Transform
Filters (LETFs).

\begin{dfn}[Linear Ensemble Transform Filter]\label{def:LETF}
A particle filter where the original sample, $\{\theta_i\}_{i=1}^M$, is transformed into a new
sample, $\{\tilde{\theta}_i\}_{i=1}^M$, according to the relation
\[
	\tilde{\theta}_j = \sum\limits_{i=1}^M \! w_{ij}\theta_i, \quad w_{ij} \geq 0, \ \sum\limits_{i=1}^M
		w_{ij} = 1,
\]
is known as a linear ensemble transform filter.
\end{dfn}

%%%%%%

\subsubsection{The Ensemble Transform Particle Filter}

The ensemble transform particle filter (ETPF) proposed by Reich in~\cite{reich2013nonparametric} is
an ensemble transform method which combines the optimal coupling ideas from
Section~\ref{sec:coupling} with the LETF framework to produce a particle filter which is consistent
in the limit of the ensemble size $M\rightarrow\infty$. The transform takes a sample of weighted
particles $\{(w_i,y_i)\}_{i=1}^M$ with law$(\mu_Y) = \pi_{Y}$ and converts it into a sample of
evenly weighted particles $\{x_i\}_{i=1}^M$ with law$(\mu_{X}) = \pi_{X}$, by means of defining a
coupling $T^*$ between the distributions $Y$ and $X$. The couplings used here are optimal in the
sense discussed in Section~\ref{sec:coupling}. The coupling $T^*$ is the solution to a linear
programming problem in $M^2$ variables with $2M-1$ constraints as defined in
\cite{reich2013nonparametric}. Choosing to maximise the correlation means that a statistic $s(X)
\rightarrow s(Y)$ as $M\rightarrow\infty$, and $s(X)$ converges faster than any other random
variable induced by a linear coupling. The density $\pi_X$ is the optimal density for approximating
$\pi_Y$ given that the mixture components have equal variance and equal weight.

\begin{table}[htpb]
\begin{algorithm}[H]
\DontPrintSemicolon
\BlankLine

\nonl \textbf{Predict}\;
Sample $\xi_i(t-1) \sim \mathcal{N}(0, \sigma^2)$.\;
Calculate $x_i^*(t) = f_{t-1}(x_i(t-1), \xi_i(t-1))$.\;
\nonl \textbf{Update}\;
For each particle calculate
\[
	w_i = \frac{\ell(D_i(t)| x_i^*(t))}{\sum_{k=1}^N \! \ell(D_k(t)| x_k^*(t))}.
\]

Find the optimal coupling,
\[
	T^* = \arg\min\limits_{T} \sum\limits_{i,j=1}^M \! t_{ij}\|x_i^*(t) - x_j^*(t)\|^2_2.
\]

Set $x(t) = M(T^*)^\top x^*(t)$.\;

\caption{The Ensemble Transform Particle Filter~\cite{reich2013nonparametric}.\label{alg:ETPF}}
\end{algorithm}
\end{table}

The ETPF is outlined in Algorithm~\ref{alg:ETPF}. It follows the same structure as the bootstrap
filter in Algorithm~\ref{alg:bs_filter}, but in this case the resampling step is performed by taking
a linear combination of the prior states rather than sampling from a multinomial distribution. The
conditions on the $t_{ij}$ are given in Equation~\eqref{eq:coupling_conditions}.

%%%%%%

\subsubsection{Approximate Multinomial Resampling}

The ETPF described in the previous section is optimal in terms of maximising the correlation between
the two marginal distributions. However the linear programming problem which must be solved for
every new empirical distribution is very costly, especially as the ensemble size grows. For the
purposes we require, discussed in Chapter~\ref{sec:PAIS}, we do not need as much accuracy as is
provided by the ETPF. For this reason we introduce a new algorithm, the approximate multinomial
resampler (AMR)~\cite{cotter2015parallel}, which shares many properties with the ETPF, but relaxes the constraint on
finding the optimal coupling. Instead we find a coupling via a greedy method in which the
correlation is high near particles with large weights, but is poor for particles with a low weight.

To perform the resampling, we split the state space into $M$, not necessarily disjoint subspaces,
$R_i\neq\emptyset$, each subspace containing at least one of the original samples $\{x_i\}$. We
define $M$ new probability distributions, $p^i$, such that $\sum_{i=1}^M \! p^i = w$, where $w$ is
the vector of the original probabilities corresponding to $x$. These probability distributions each
satisfy the conditions
\[
	\tilde{p}_k^i \leq w_k, \quad \|\tilde{p}^i\|_1 = \frac{1}{M},\quad \text{and}\quad p^i =
		M\tilde{p}^i.
\]
At this point, we could still be talking about the ETPF. The distinction comes in the way that we
choose the $R_i$ and $\tilde{p}^i_k$. We choose to first pick the particle, $x_J$, with the highest
weight and then expand a ball around it until the ball contains $1/M$ of the total probability mass.
The particle furthest from $x_J$ inside the ball may have its weight reduced by the amount required
for $\|\tilde{p}^i\|_1=1$. All other particles in the ball will have $\tilde{p}^i_k = w_k$. Once
this is done, particles which have had all their probability mass assigned to a distribution $p^i$
are removed, and we choose the particle with the next highest weight.

This method ensures that areas with a high weight have a particle chosen in that area again. The
greedy nature of the algorithm means that the correlation is sub optimal. In Algorithm~\ref{alg:AMR}
we present the pseudo-code form of this algorithm.

\begin{table}[htpb]
\begin{algorithm}[H]
\DontPrintSemicolon
\BlankLine

Obtain a weighted sample $\{(w_i, y_i)\}_{i=1}^M$.\;
\For{$i=1,\dots,M$}{
	$J = \arg\max_j \! w_j$.\;
	$p_{J,i} = 1/M \wedge w_J$.\;
	$w_J = w_J - p_{J,i}$.\;
	\While{$\sum_j p_{j,i} < 1/M$}{
		$K = \arg\min_{k\in\{k|w_k>0\}} \! \|y_J - y_k\|$.\;
		$p_{K,i} = (1/M-\sum_j \! p_{j,i}) \wedge w_K$.\;
		$w_K = W_K - p_{K,i}$.\;
	}
}
$x = MP^\top y$.\label{algline:amr_mean}\;
\caption{The Approximate Multinomial Resampler~\cite{cotter2015parallel}.\label{alg:AMR}}
\end{algorithm}
\end{table}

In Line~\ref{algline:amr_mean} of Algorithm~\ref{alg:AMR} we see that the output of the algorithm,
an evenly weighted sample, takes the form required by Definition~\ref{def:LETF} to be an LETF. We
note that the AMR algorithm, like the ETPF algorithm preserves the mean of the sample,
\[
	\hat{\mu}_X = \frac{1}{M}\sum_i \! x_i = \frac{1}{M}\sum_i \sum_k \! p_{k,i}y_k = \frac{1}{M}\sum_k
		w_ky_k = \hat{\mu}_Y.
\]
Here we use that the columns of $P$ are the probability distributions $p^i$, and the sum of these
distributions is equal to the weights, $w$, hence
\[
	\sum_i \! p_{k,i}y_k = (p^1_k + \dots + p^M_k)y_k = w_ky_k.
\]

%%%%%%

\subsubsection{Numerical Comparison of the Filter Resamplers}

We have designed the AMR with the aims of being numerically cheaper than the ETPF, and more accurate
than the BSR. Therefore we now present numerical examples which demonstrate that we have achieved
this. To compare the algorithms, we write a simple MC algorithm which compares the ability of each
algorithm to preserve the statistics of a weighted sample in an evenly weighted sample.

\begin{table}[htpb]
\begin{algorithm}[H]
\DontPrintSemicolon
\BlankLine
\For{$k=1,2,\ldots,\frac{N}{M}$}{
	Sample $y_j \sim Q$, for $j = 1, \ldots, M$.\;
	Weight $y_j$ under $F$,
	\[
		w_j = F(y_j) / Q(y_j).
	\]

	Resample 
	\[
		\hat{F}_k = \sum_j w_j\delta_{y_j}(\cdot) \rightarrow \sum_j n_j\delta_{x_j}(\cdot) = \hat{G}_k \approx \hat{F}_k.
	\]

	Calculate error in the statistic
	\[
		s_k = \frac{|s(\{x_j\}) - s(\{(w_j, y_j)\})|}{|s(\{x_j\})|}.
	\]
}
Return mean($\{s_k\}$).\;
\caption{Comparing Resamplers.\label{alg:ETMC}}
\end{algorithm}
\end{table}

In Algorithm~\ref{alg:ETMC} we create random empirical weighted distributions, $\hat{F}$, which are an
approximation of a chosen distribution $F$. We then resample from $\hat{F}$ to create a new
distribution $\hat{G}$. The new empirical distribution, $\hat{G}$, follows roughly the same
distribution but some error is introduced since we require an integer number,
$n_j$, of each particle. We want to calculate how big the error is between the distributions of $\hat{F}$ and
$\hat{G}$. To do this we calculate some chosen statistics of both $\hat{F}$ and $\hat{G}$ and
calculate the relative difference between these statistics. This is done for each resampler with a fixed total sample size of $N$,
and we vary $M$ to see how the relative error varies with the ensemble sizes.

\begin{figure}[htb]
\centering
\subfigure[Relative error in $\mathbb{E}(X)$]{\includegraphics[width=0.45\textwidth]{"images/AMR_EX"}}
\subfigure[Relative error in
$\mathbb{E}(X^2)$]{\includegraphics[width=0.45\textwidth]{"images/AMR_EX2"}}\\
\subfigure[Relative error in $\mathbb{E}(X^3)$]{\includegraphics[width=0.45\textwidth]{"images/AMR_EX3"}}
\subfigure[Cost of resamplers per iteration]{\includegraphics[width=0.45\textwidth]{"images/AMR_speed"}}\\
\caption{Relative errors representing the error between $\hat{F}$ and $\hat{G}$ as described in Algorithm~\ref{alg:ETMC}, where $\hat{G}$ is constructed using the Bootstrap, ETPF and AMR algorithms.}
\label{fig:AMR}
\end{figure}

To test the resamplers we chose a proposal distribution, $Q = \mathcal{N}(1,2)$, and a target
distribution $F = \mathcal{N}(2,3)$. Figure \ref{fig:AMR} (a)-(c) shows how the relative errors in
the first three raw moments varies with ensemble size $M$. As expected, the AMR lies somewhere
between the high accuracy of the ETPF and the less accurate BSR. Note that only the error for the
BSR is presented for the first moment since both the ETPF and the AMR are mean preserving. Figure
\ref{fig:AMR} (d) shows how the computational cost, measured in seconds, scales with the ensemble
size. These results demonstrate that the AMR behaves how we wish, and importantly ensures that
exactly one sample of the output will lie in each region with higher than average density.

%%%%%%

\subsubsection{Scaling of Resampling Algorithms with Dimension}
Ideally the dimension of particles in a filter will not affect the convergence of the method.
Unfortunately, the approximation of continuous density functions is computationally infeasible in
high dimensions with the ensemble size required to achieve convergence growing exponentially in the
dimension\cite{silverman1986density,snyder2008obstacles}.

\begin{figure}[htb]
\begin{center}
\includegraphics[width=0.7\textwidth]{images/ET_dim_scaling}
\caption{Demonstration of the ETMC algorithm rapidly failing as the dimension increases in problem
$G_1$ defined in Chapter~\ref{sec:PAIS}. The relative error shown is that between $\mathbb{E}_{\hat{F}}[X]$ and $\mathbb{E}_{\hat{G}}[X]$.}
\label{fig:ET_dim_scaling}
\end{center}
\end{figure}

Figure~\ref{fig:ET_dim_scaling} shows how quickly the ETMC algorithm loses its ability to accurately
represent the mean of the target distribution as the dimension increases. Clearly for a simple
Gaussian problem, we can not resample in much more than 10 dimensions and hope to recover a
representative sample without massively increasing the ensemble size. This phenomenon is known as
the curse of dimensionality. The AMR and BSR suffer in a similar manner as the dimension increases.

%%%%%%%%%%%%

\subsection{Couplings between Arbitrary Probability Measures}\label{sec:intro_TM}

In the previous sections we have looked at couplings which convert a weighted sample into an
unweighted sample while preserving the approximate empirical distribution. In this section we look at
how we can couple two arbitrary measures. In the following discussion we concentrate on designing a
coupling which maps from an arbitrary probability distribution onto a standard Gaussian
distribution. The main difference here is that we would like to construct a map between the
continuous distribution $F$ and a \emph{reference measure}, $\mathcal{N}(0, \text{I})$, rather than
between the approximation of $F$, $\hat{F}$, and the approximation of $\hat{F}$, $\hat{G}$.

Obviously constructing an exact map for arbitrary $F$ is infeasible, so we must again use an
empirical distribution $\hat{F}$ where $\hat{F}\rightarrow F$ in the limit of infinite samples,
however we can ensure that we are coupling $\hat{F}$ to the continuous distribution $\mathcal{N}(0,
1)$. This convergence must be achieved by continual sampling from $F$, which for an arbitrary
distribution can be expensive. To speed up this process~\cite{parno2014transport} proposes an MCMC
algorithm which sequentially updates the distribution $\hat{F}$ and recalculates the map so that
after each update, sampling from $F$ becomes cheaper. Calculating the transport map between
$\hat{F}$ and $\mathcal{N}(0, \text{I})$ allows us to sample more efficiently because it reduces the
sampling problem to sampling from a standard Gaussian distribution and then mapping these Gaussian
random variables onto an arbitrary probability measure. Which for the class of maps considered is
very cheap. These couplings are discussed further in Chapter~\ref{sec:map}.

@@

So far we have seen that the PAIS algorithm is efficient for low dimensional posterior sampling.
PAIS is particularly useful if the posterior is multimodal, or inconveniently shaped such as the
Rosenbrock density discussed in Chapter~\ref{chapter:HMC}. However, we have had trouble when
attempting to increase the dimension of the parameter space. This was seen when we attempted to
recover the initial condition for the wave equation in Chapter~\ref{sec:PAIS}.

In this Chapter we look at a method which will allow us to improve efficiency in low dimensions, and
will also allow us to sample more efficiently in higher dimensions. The ideal sampling
situation in MCMC is to explore an $n$-dimensional Gaussian distribution (or other standard distribution) using a near proportional
proposal distribution. We therefore aim to transform an arbitrary distribution, $\mu_\theta$, in to a Gaussian distribution, which we can then sample from efficiently, and finally transform this sample back into a sample from $\mu_\theta$. In this way we hope to be able to sample from any distribution $\mu_\theta$ with a similar efficiency as we obtain when sampling from a Gaussian distribution. In particular for the PAIS algorithm, resampling schemes are more efficient in higher
dimensions if the dimensions are uncorrelated. The work of Moselhy and
Marzouk~\cite{el2012bayesian}, Parno and Marzouk~\cite{parno2014transport} and
Parno~\cite{parno2015transport} provides a method of constructing such a map as part of an adaptive
Metropolis-Hastings algorithm.

%Given some constraints it is computationally tractable to adaptively construct invertible maps
%during an MCMC simulation. These maps construct a continuous coupling between a target and
%reference distribution, where the latter is chosen to be a standardised Gaussian. How close this reference
%distribution is to Gaussian depends on three considerations; firstly the constraints applied. These
%constraints, discussed later, ensure that the map exists, is unique, and is invertible. The
%necessity for the map to be invertible, and preferrably easily invertible, restricts the space of
%possible couplings to the point where it is unlikely that the exact map is included in our space of
%feasible transference plans. Secondly, for our purposes, the map takes the form of a polynomial
%chaos expansion which requires infinitely many terms to exactly represent the coupling. This is of course not possible on current hardware. Finally we
%will be constructing this map using an optimisation problem where the objective function depends on
%the sample path of our MCMC algorithm. This sample does not represent the target exactly until $N\rightarrow\infty$, and so the expectation is only calculated approximately.
%
%While the restrictions on the map are essential for well-posedness, the approximate map which
%results from these constraints can be made arbitrarily accurate by increasing the number of basis
%functions used in the expansion, and increasing the number of samples used in the optimisation
%problem. In practise it appears that we need not aim to achieve a high accuracy solution. Excellent
%results have been observed using monomial basis functions of total order 3, and we need a sample of
%size of only $\mathcal{O}(10^3)$ to obtain a well converged map.
